Combinatorial Mixture Models for Single Cell Assays

### 2012-04-04 MIMOSA.tex
Restructuring..
ongoing.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\begin{document}
\maketitle
\begin{abstract}
Immunological endpoints in vaccine trials are measured through a number of different assays that often provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific cell sub--populations. While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. For example, in flow cytometry, individual cells are often classified as either positive or negative for a marker based on a predetermined threshold.  One such assay is the intracellular cytokine staining assay that is used to assess an individual's immune response to a vaccine by measuring the abundance antigen--specific T--cell subpopulations producing different cytokines. Cells are classified as cytokine positive or negative and the resulting cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells. ICS assays are generally used to identify ``vaccine responders''. These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline. The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern. The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons. It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test. Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.  In this paper we present a coherent framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from single--cell assays.  Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously, sharing information across observations using empirical Bayes methods. The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently. Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test. 
\end{abstract}

\section{Introduction}
Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a homogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

A motivating example from vaccine trial research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells. Not all T--cells can recognize all antigens. Those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines that further promote the immune response. After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future. None the less, these antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells. The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.

@@

We develop our framework in the context of ICS data. While there is no standard approach to analyzing ICS assay data, Fisher's exact test, as well as more ad--hoc methods, are commonly applied on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa}. Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}. In the context of single--cell assay data the assumption of fixed margins is not satisfied since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect. More importantly, Fisher's exact test does not share information across observations, resulting in low power when counts are small.

The framework developed in this paper addresses these issues explicitly. We present a combinatorial mixture model framework for the analysis of single--cell assay data in which multivariate measurement are made individual cells. The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.

\section*{Materials and Methods}

\subsection*{Vaccine Trial ICS Dataset Description}
HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}. The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}. Vaccine was given at two increasing doses, as well as a placebo. T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}. The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}. The sample size consisted of 20 vaccine and four placebo recipients. Statistical analysis of the original positivity calls is described in the original publication~\cite{Peiperl:2010ej}.

\subsection*{Data Import, Preprocessing, and Gating}
The gated ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4). Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.

\subsection*{Statistical Analysis of Responder and Non--responder calls}
Below, we summarize the methods for statistical analysis of responder and non--responder calls in the published trial, as well as the methods compared in this paper.
\subsubsection*{Statistical Analysis in the Published Trial}
The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}. In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual. In the original trial, significance was measured via one--sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual. A discrete Bonferroni adjustment for multiple comparisons was applied, and stimulations with an adjusted p--value below $\alpha = 0.00001$ were called positive.

\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.

\subsection*{Two Competing Beta--Binomial Models}
Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach. We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples). For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively. Note that usually, $N_u \ne N_s$. The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants. For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table: 
% latex table generated in R 2.14.0 by xtable 1.5-6 package
% Wed Oct 12 09:41:50 2011
\begin{table}[ht]
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for cytokine positive and cytokine negative events between stimulated and unstimulated conditions}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}

The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as:
\begin{align}
  \text{if }&p_{si}=p_{ui}\equiv p_{0i};& n_{si} \sim \mathrm{Bin}(N_{si},p_u);\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null}\\
 \text{if }&p_{si}>p_{ui};& n_{si} \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:alternate}
 \end{align}
Where $p_{si}$ and $p_{ui}$ are the unobserved proportions. Equation~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control. Equation~\eqref{eq:alternate} represents the  alternate hypothesis where the cytokine response is stronger in the stimulation than in the control. We place a common Beta prior on the $p_{si}$ and $p_{ui}$ across individuals, as shown:

 \begin{align}
 p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s)\label{eq:stimprior}\\
 p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:nullprior}
 \end{align}
If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models. Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
 \begin{align}
  	\mathrm{Pr}(y_i|\alpha_u,\beta_u)
	&=\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}\\
	\intertext{with marginal log--likelihood:}
	\begin{split}
	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
	\end{split}
 \end{align} 
Thus, in the case of no response to stimulation, the counts for the stimulated and unstimulated samples are modelled as draws from the same "unstimulated" Beta--binomial distribution. Note that the unobserved parameters, $p_{si}, p_{ui}$ have been integrated out to give the marginal log--likelihood.

If the data is generated by model~\eqref{eq:alternate}, the posterior probability of the data is given by:
\begin{align}
	\begin{split}
\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
	 \frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
\label{eq:model2post}
\end{split}
	\intertext{with marginal log--likelihood:}
\begin{split}
\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right. \\ \left. \log\left(\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)\right) + \right. \\ \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right) \right. \right. \\  \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)}\left(\mathrm{I_{1-p_{ui}}}(N_{si}-n_{si}+\beta_s,n_{si}+\alpha_s)\right)d p_{ui}\right] \\ -  \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right) \right. \right. \\  \left. \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)} \left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}\right] \right\}\label{eq:model2MLL}
\end{split}
\end{align}

The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions. We call this the \emph{constrained} model.

The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of Beta distribution with parameters $\alpha_s,\beta_s$, leaving a 1--dimensional integration for the ratio of normalizing constants.

Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.

>>>>

\begin{document}
\maketitle

\begin{abstract}
Immunological endpoints in vaccine trials are measured through a variety of assays that  provide single--cell measurements of multiple genes and proteins in specific immunological cell populations. %intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific immunological cell populations. 
Using single--cell data, we consider the problem of identifying subjects where these cell populations exhibit differential responses under different experimental conditions.  
%While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. 
For example, in the intracellular cytokine staining assay from flow cytometry, individual cells are classified as either positive or negative for a marker based on a predetermined threshold.  %One such assay is the intracellular cytokine staining assay that
The assay is used to assess an individual's immune response to a vaccine by measuring the number of antigen--specific cells  producing different cytokines in different T--cell subpopulations in response to different antigen stimulations.  Individuals whose T--cells exhibit increased production of a cytokine in response to stimulation are termed ``positive'' for that cytokine, and multiple such ``positivity calls'' are used to identify vaccine responders. 
%Cells are classified as cytokine positive or negative and the resulting The cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells. The ICS assay is generally used to identify ``vaccine responders''. These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline. The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern. The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons. It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test. Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.  
Here we present a framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from such single--cell assays.  Our method models cellular responses in a marker--specific manner, treating the responding and non--responding observations as separate components in the model. Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls. We compare our method against Fisher's exact test and show how it can be extended to model multivariate (multiple markers) cellular responses. In simulations and in HIV vaccine trial data we find that our method has higher sensitivity and specificity than Fisher's exact test for positivity calls. 
% simultaneously, sharing information across observations using empirical Bayes methods. The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently. Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test. 
\end{abstract}

\section{Introduction}
Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells. While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future\cite{McKinstry:2010ei}. These antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells. The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw, Bendall:2011wf}.

@@

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.

%on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}. In the context of single--cell assay data the assumption of fixed margins is not satisfied since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect. More importantly, Fisher's exact test does not share information across observations, resulting in low power when counts are small.

The framework developed in this paper addresses these issues explicitly. We present a combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells. The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells. Importantly, our approach shares information across observations by means of prior distributions on the responding, stimulated samples and the non--responding samples thereby increasing sensitivity and specificity to detect responding cell populations. The cell counts from the stimulated and unstimulated experiments are modelled independently, and different combinations of markers are represented as different mixture components in the framework. This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines). This is a flexible approach that avoids the some of the drawbacks of global tests when only a few of the many measured markers show differences~\cite{Nason:2006dx}.

\section{Data structure and notation} 
We analyzed two data sets from two different immunological single--cell assays typically used in vaccine trials, one flow cytometry data set, and one single--cell gene expression data set.

\textit{Flow cytometry}: This ICS data is from a phase--I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post--vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder and non--responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess response rates of CD4 and CD8 T--cell populations to different antigens.

\textit{Fluidigm single--cell gene expression}: This is a single--cell gene expression data set of flow--sorted CD8 T--cells from sixteen individuals. The T--cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single--cell level and compared to paired, unstimulated controls.

In the remainder of the paper, we use the following notation to describe our model. We observe cell counts from $P$ subjects in two conditions (stimulated and unstimulated). Each cell can either be positive or negative for a marker. We let $i$ index the subjects and $k$ index the cytokines. The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively. The total number of cells is denoted by $N$, and given in our model, while $n$, the number of positive cells is part of the observe data. Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive. Similarly, $N_{ui}$, $n_{ui}$ represent the total, and cytokine positive cell counts in the unstimulated condition for the same subject. For each cytokine and each stimulation, given the total cell counts, the observed data is a matrix of size $2 \times P$, where the $i$'th row is a subject, $\mathbf{y}_i = <n_{si},n_{ui}>$.

\section{Differential expression with one marker}
The data for a single marker can be represented in contingency table form:

\begin{table}[ht]
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for cytokine positive and cytokine negative events between stimulated and unstimulated conditions}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}
We consider a marker to be differentially expressed between conditions if the number of cells in the stimulated samples is significantly different from the number of cells in the control sample. The model is described below.

\subsection{Beta--Binomial Model for One Marker}
The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as  binomial random variables.
\begin{align}
(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non--responder}\\
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
 \end{align}

Where $p_{si}$, $p_{ui}$ are the unobserved proportions. Model~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control, and thus no difference between the proportion of cytokine positive cells in stimulated and control samples. Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in cytokine response between the stimulation and the control, and thus a difference in the proportion of cells between stimulated and control samples. In the case of no response, the counts for the stimulated and unstimulated samples are independent draws from the same  binomial distribution.

\subsection{Prior}
Our model shares information across responders and non--responders, respectively. This is done by placing Beta priors on the $p_{si}$, $p_{ui}$, across subjects:

 \begin{align}
 p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s);&
 p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior}
 \end{align}

Additional sharing of information is done through the $\alpha_u, \beta_u$ hyper--parameters, which are common for the null and alternative models. 
%If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models. Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
\subsection{Posetrior}
\subsubsection{Null Model}
Given the above formulation, the posterior for the null model is Beta--binomial, with marginal likelihood (integrating out $p_u$):

 \begin{align}
  	\mathrm{L}(\alpha_u,\beta_u|\mathbf{y})
	&=\prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
%	\intertext{Integrating out the proportions gives the marginal log--likelihood.}
%	\begin{split}
%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
%	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
%	\end{split}
 \end{align} 
 \subsubsection{Alternative Model - Unconstrained}
 If the data is generated by model~\eqref{eq:alternate} then we have  $p_s\ne p_u$, and the marginal likelihood of the alternative is:
\begin{align}
	\begin{split}
\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\label{model2:unconstrained}
\end{split}
\end{align}

\subsubsection{Alternative Model - Constrained}
For ICS data, however, we are generally interested in the one--sided case where $p_s>p_u$, which is consistent with a positive response to stimulation (i.e., more cells are producing the marker after stimulation). In this constrained case, integrating out $p_s$, $p_u$ leads to an additional factor, $C$, in the marginal likelihood, which is just the ratio of normalizing constants for the prior and the likelihood with constraints:
\begin{align}
\begin{split}
	 C=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
\label{eq:model2post}
\end{split}
%	\intertext{with marginal log--likelihood:}
%\begin{split}
%\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right. \\ \left. \log\left(\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)\right) + \right. \\ \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right) \right. \right. \\  \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)}\left(\mathrm{I_{1-p_{ui}}}(N_{si}-n_{si}+\beta_s,n_{si}+\alpha_s)\right)d p_{ui}\right] \\ -  \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right) \right. \right. \\  \left. \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)} \left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}\right] \right\}\label{eq:model2MLL}
%\end{split}
\end{align}

%The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions. We call this the \emph{constrained} model.

The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional numerical integration for calculating the ratio of normalizing constants.

%Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.

\subsection{The Mixture Model Framework}

The two models (null and alternative), are represented as two components in our mixture model framework. The observed data is thus a mixture of observations drawn from one or the other of these components (i.e., each subject is either a responder, or a non--responder). Setting  $\theta_1=\{\alpha_u,\beta_u\},\theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, the likelihood for the mixture model is:

%\begin{align}
%f(y_i|\alpha_{u},\beta{u},\alpha_{s},\beta_{s}) = (1-\pi)f_1(y_i|\alpha_u,\beta_u) + \pi f_2(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)
%\end{align}
